#Impact of wheather events on USA citizens

The data analysed here contains information about impact of whether events on US citizens. The impact is divided into two major groups:

* personal (fatalities and injuries)
* financial (property damage and crops damage).

The analysis below will show that "Heavy surf and wind" event is the most damaging one in both above categories.

###Data Processing
DPLYR and GGPLOT2 libraries will be used to process data and prepare results.
```{r}
library(dplyr)
library(ggplot2)
```
Data for producing the report is organized in folders.

* code - this folder will contain all code files containing routines run on the data
* data - this folder will contain raw and processed data
* documentation - this folder will contain R markdown files for assembling the report as well as the final HTML/PDF report
* figure - this folder will contain graphical images containing figures (plots) included in the report

The raw data is downloaded from [Coursera](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) website repository for this particular course and then stored in the project folder under the "data" folder.
As a first step, the data is read into memory to perform basic exploratory analysis:
```{r cache=TRUE}
rawData <- read.csv("../data/repdata-data-StormData.csv.bz2")
colnames(rawData)
head(rawData)
summary(rawData)
```

###Results

####Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
In order to determine which types of events (EVTYPE) are most harmful for the population health across United States, a smaller data frame containing only 3 columns will be extracted:

* EVTYPE
* FATALITIES
* INJURIES

This will then be grouped by event type and ordered (descending).
The result will allow to view the most harmful events at the top of the table.
```{r}
healthImpact <- rawData[c("EVTYPE", "FATALITIES", "INJURIES")]
evTypesSummary <- data.frame(unique(healthImpact$EVTYPE))
colnames(evTypesSummary) <- c("EVTYPE")
evTypesSummary["SUM_FATALITIES"] <- tapply(healthImpact$FATALITIES, healthImpact$EVTYPE, sum)
evTypesSummary["SUM_INJURIES"] <- tapply(healthImpact$INJURIES, healthImpact$EVTYPE, sum)
evTypesSummary <- arrange(evTypesSummary, desc(SUM_FATALITIES), desc(SUM_INJURIES))
head(evTypesSummary, n = 20)
```
As per the results in the table, it's apparent that the biggest whether event in terms of fatalities and injuries in USA is **"Heavy surf and wind"**, followed by **"FLOOD/RAIN/WINDS"**.

To visualize the ratios a bar chart has been generated:
```{r}
plotData <- head(evTypesSummary, n = 15)
tmpInjuriesDmg <- select(plotData, EVTYPE, SUM_INJURIES)
colnames(tmpInjuriesDmg) <- c("EVTYPE", "DMG")
tmpInjuriesDmg["DMG_TYPE"] <- "Injuries"
plotData <- plotData[c("EVTYPE", "SUM_FATALITIES")]
colnames(plotData) <- c("EVTYPE", "DMG")
plotData["DMG_TYPE"] <- "Fatalities"
plotData <- rbind(plotData, tmpInjuriesDmg)

plotData$EVTYPE <- factor(plotData$EVTYPE, levels = plotData$EVTYPE[1:15])

options(scipen=1000000)
g <- ggplot(plotData, aes(x = EVTYPE, y = DMG, fill = DMG_TYPE))
g <- g + geom_bar(stat="identity") + xlab("Event Type") + ggtitle("Relative health impact")
g <- g + ylab("Count") + guides(fill=guide_legend(title=NULL))
g <- g + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0))
g

```


####Across the United States, which types of events have the greatest economic consequences?
In order to determine which types of events have most significant economic impact on US citizens, another smaller data set will be extracted from the "rawData" data set containing:

* EVTYPE
* PROPDMG
* CROPDMG

This will then be grouped by event type and ordered (descending). Only 15 most impactful events will be listed.
```{r}
economicImpact <- rawData[c("EVTYPE", "PROPDMG", "CROPDMG")]
economicImpact <- arrange(economicImpact, desc(PROPDMG, desc(CROPDMG)))
evTypesSummary <- data.frame(unique(economicImpact$EVTYPE))
colnames(evTypesSummary) <- c("EVTYPE")
evTypesSummary["SUM_PROPDMG"] <- tapply(economicImpact$PROPDMG, economicImpact$EVTYPE, sum)
evTypesSummary["SUM_CROPDMG"] <- tapply(economicImpact$CROPDMG, economicImpact$EVTYPE, sum)
evTypesSummary["SUM"] <- evTypesSummary["SUM_PROPDMG"] + evTypesSummary["SUM_CROPDMG"]
evTypesSummary <- arrange(evTypesSummary, desc(SUM))
head(evTypesSummary, n = 15)
```

As per the results in the table, it's apparent that the biggest whether event in terms of damages to properties and crops USA is **"Blowing Snow"**, followed by **"Beach Erosion"** and **"Wintry mix"**.

To visualize the ratios a bar chart has been generated:
```{r}
plotData <- head(evTypesSummary, n = 15)
tmpCropDmg <- select(plotData, EVTYPE, SUM_CROPDMG)
colnames(tmpCropDmg) <- c("EVTYPE", "DMG")
tmpCropDmg["DMG_TYPE"] <- "Crop"
plotData <- plotData[c("EVTYPE", "SUM_PROPDMG")]
colnames(plotData) <- c("EVTYPE", "DMG")
plotData["DMG_TYPE"] <- "Property"
plotData <- rbind(plotData, tmpCropDmg)

plotData$EVTYPE <- factor(plotData$EVTYPE, levels = plotData$EVTYPE[1:15])

options(scipen=1000000)
g <- ggplot(plotData, aes(x = EVTYPE, y = DMG, fill = DMG_TYPE))
g <- g + geom_bar(stat="identity") + xlab("Event Type") + ggtitle("Relative economic impact")
g <- g + ylab("Damage") + guides(fill=guide_legend(title=NULL))
g <- g + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0))
g

```
